FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-09-16 manuscript.Rmd
update 2024-09-16 09:57:53.8483

The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
Since they were first introduced, scholars have demonstrated their relevance for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
To this end, we introduce the concept of "FAIR theory",
a digital representation of theory compliant with the FAIR principles.

@@

There are clear indications that psychological theory has limited interoperability. 
For one, theories are rarely *"refuted nor corroborated, but instead merely fade away as people lose interest"* [@meehlTheoreticalRisksTabular1978].
There even appears to be a norm *against* theory interoperability, as evident from the quip that 
*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008]
To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
For example, it should be possible to integrate theory directly into analysis workflows:
to derive hypotheses, select control variables, and guide model specification.
A lack of formalization [REF Robinaugh] and ambiguity [REF Frankenhuis] prevent interoperability in such a practical sense. 
Additionally, theories should be interoperable with each other:
for instance, it should be possible to embed a specific theory about the process of emotion regulation [REF Gross] within a theory of emotion regulation development [REF Morris].

<!-- The ultimate goal of FAIR is to optimise the reuse of data. To achieve this, metadata and data should be well-described so that they can be replicated and/or combined in different settings. -->
Is it reusable? Scholars would rather invent something new than reuse somebody else's toothbrush (which is antithetical to the empirical cycle).

@@

@@

@@

This protocol shows how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG. We could use that to illustrate updating FAIR theory:

>>>>

The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
Since their inception, scholars have demonstrated their relevance for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
To this end, we introduce "FAIR theory":
a digital representation of theory compliant with the FAIR principles.
FAIR Theory has the potential to substantially advance the efficiency of scholarly communication and accelerate cumulative knowledge acquisition.

@@

There are clear indications that psychological theory has limited interoperability. 
For one, theories are rarely *"refuted nor corroborated, but instead merely fade away as people lose interest"* [@meehlTheoreticalRisksTabular1978].
To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
For example, it should be possible to integrate theory directly into analysis workflows:
to derive hypotheses, select control variables, and guide model specification.
The aforementioned lack of formalization [REF Robinaugh] and ambiguity [REF Frankenhuis] prevent interoperability in such a practical sense.
Additionally, theories should be interoperable with each other:
for instance, it should be possible to embed a specific theory about the process of emotion regulation [REF Gross] within a theory of emotion regulation development [REF Morris].

The ultimate goal of the FAIR principles is to facilitate the reuse digital research artefacts.
While theory is ostensibly held in high regard in the social sciences,
there also appears to be a norm *against* theory reuse.
This is evident from the quip that 
*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
In our workshops on FAIR Theory, we often hear questions such as "who owns theory",
and "who determines how a theory may be reused or changed"?
Licensing theories for reuse provides an unambiguous answer to such questions.

One important consideration is that there may be legal limits to the rights authors can claim over theory.
An important concept in copyright law is the *idea-expression dichotomy* [REF].

o. Ideas, facts, and concepts are not protected by  Although they are not protectable by copyright, the expression of those ideas, facts, and concepts are protectable, such as in a description, explanation, or illustration or as a database of facts. Separating an idea from the expression or manifestation of that idea is known in copyright law as the 
Copyright explicitly does not
*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*,
but it may extend to creative works expressing that idea (e.g., writing, visual illustrations).
Copyright pertaining to recipes offer an interesting perspective:
Whereas the recipes themselves are not protected, *"substantial literary expression in the form of an explanation or directions"* can be protected by copyright.
Extrapolating to FAIR Theory, it would seem that vague, ambiguous verbal explanations of theories are more likely to qualify for copyright protection than formal theories.
If copyright does not cover (formal) theory,
then it would be pointless to adopt a license that assumes copyright protection.
Furtermore, even if copyright would apply, research is covered under "fair use" excemptions to copyright law, so copyright would impose few restrictions to theory reuse in scholarly communication.
In sum, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
it explicitly waives all rights and encourages reuse.
Simultaneously, to comply with the interoperability principle, the principle of comprehensive citation [REF TOP guidelines],
and the spirit of good scholarship,
it is essential that scholars do comprehensively cite theory,
including prior work that new theories are based on.

## FAIR Theory and Recognition & Rewards

@@

## Version Control

* One source of potential improvements of theory methodology that has not been previously considered is computer science.
* The process of "iteratively improving" digital objects - in this case, computer code - is well understood.
* Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
* This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
* We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories
* FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner. Version control also enables diffuse collaboration in theory development, as other researchers can submit "pull requests" to suggest modifications of a theory, or can "fork" existing theories to create a spin-off from an existing theory.

@@

## Formalizing the Empirical Cycle

In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
As several authors have taken inspiration from the work by De Groot,
we compare our interpretation of the original theory to the interpretation of others.
Originally, the theory has the following structure:

```

digraph {

  observation;
  induction;
  deduction;
  test;
  evaluation;

  observation -> induction;
  induction -> deduction;
  deduction -> test;
  test -> evaluation;
  evaluation -> observation;

}
```

Subsequently, Wagenmakers and colleagues modified the theory by *"[adding the] Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"*:

```
digraph {

  subgraph cluster_discovery {
    label="Discovery";
    hypothesis [label="New hypothesis"];
    prediction [label="New prediction"];
  }
  data  [label="Old knowledge and old data"];      
  subgraph cluster_justification {
    label="Justification";
    test [label="Test on new data"];
    evaluation;
  }

  data -> hypothesis [label="Speculate & explore"];
  hypothesis -> prediction  [label="Deduce"];
  prediction -> test  [label="Design new experiment"];
  test -> evaluation  [label="Statistical analysis"];
  evaluation -> data  [label="Knowledge accumulation"];

}
```

Note, however, that there appear to be further changes:
the phases of the cycle have been renamed,
and the annotations suggest a move towards experimental empirical psychology that was absent in the original formulation.
Moreover, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.

Our work, too is inspired by De Groot, but our take on the empirical cycle is different again:

```
digraph {

  theory;
  prediction;
  test [label="inferential procedure"];
  observation;

  theory -> prediction [label="deduction"];
  prediction -> test;
  test -> observation;
  observation -> theory [label="generalization"];

}
```

In our representation,
induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory.
For example, the refutation of a hypothesized effect,
or the serendipitous observation of some pattern in data,
might be a reason to revise or construct theory.
Induction, incidentally, also occurs within the link from prediction to testing:
in the form of the inductive bias of methods used to perform the test,
and auxiliary assumptions that must be made to address remaining theoretical ambiguities.

<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced. -->
<!-- ``` -->

@@

Some have argued that *causal explanations* are a property of good theory [REF Meehl, etc?].
According to Pearl and colleagues,
explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
and could even be integrated into the analysis environment.

In this example, we illustrate how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG. We could use that to illustrate updating FAIR theory:

### 2024-09-16 manuscript.tex
update 2024-09-16 09:57:53.8483

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 1808}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# 50b18562.

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable {[}REF{]}.
Since they were first introduced, scholars have demonstrated their relevance for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
To this end, we introduce the concept of ``FAIR theory'',
a digital representation of theory compliant with the FAIR principles.

@@

While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
we provide a narrative review of examples and counterexamples for each criterion.
One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
The modular publishing movement may ameliorate the Findability of psychological theory.
In this approach, distinct resources are individually published as citable academic output.
Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
Some have sought to address limitations of theories'findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website.
Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).

With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
Open Access publishing is a practice that increases the accessibility of all academic output, including theory.
Another factor curtailing the accessibility of theory is ambiguity;
a lack of formalization introduces a dependency on the original author for clarification.
The discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
Dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
This also relates to the problem of translation {[}REF Duhem{]}:
it is not possible to entirely formalize an idea to enable unambiguous interpretation.
Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.

Is it interoperable? Theories rarely change,

Is it reusable? Scholars would rather invent something new than reuse somebody else's toothbrush (which is antithetical to the empirical cycle).

@@

@@

@@

This protocol shows how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG. We could use that to illustrate updating FAIR theory:

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 2657}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# d1a82fc7.

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable {[}REF{]}.
Since their inception, scholars have demonstrated their relevance for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
To this end, we introduce ``FAIR theory'':
a digital representation of theory compliant with the FAIR principles.
FAIR Theory has the potential to substantially advance the efficiency of scholarly communication and accelerate cumulative knowledge acquisition.

@@

While a comprehensive analysis of the present state of FAIR Theory in psychology is beyond the scope of the present paper,
we provide a narrative review of examples and counterexamples for each criterion.

One factor that contributes to poor findability of psychological theory is that the primary unit of dissemination and search in psychology is the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
or even an agreed-upon keyword (model, framework, etc are often used interchangeably with theory).
Modular publishing could ameliorate the Findability of psychological theory.
In this approach, distinct resources are individually published as citable academic output.
Such output can still be linked; for example, a FAIR theory could have an accompanying paper.
Another effort to improve theories' findability is post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website.
Similarly, Borsboom and colleagues seek to establish a dictionary of psychological ``phenomena'' (which are not strictly theories, but are patterns reliably evidenced by data that theory should seek to explain).

With regard to Accessibility, publishing behind paywalls certainly has a negative impact.
Open Access publishing increases the accessibility of all academic output, including theory.
Another factor curtailing the accessibility of theory is ambiguity;
a lack of theory formalization introduces a dependency on the original author for clarification.
The discourse on ``Great Man Theorizing'' touches upon the problems this introduces {[}REF Guest et al moral theory{]}.
For example, dependency on interpretation by the author creates a potential for gatekeeping - the author could insist that work requires their involvement, which violates checks and balances of scientific research.
Moreover, if a theory is refuted, its author could claim that the author of the refuting paper did not interpret the theory correctly.
To some extent, this relates to the problem of translation {[}REF Duhem{]}:
it is not possible to entirely formalize an idea to enable unambiguous interpretation.
Nonetheless, taking care to formalize a theory to the maximum extent possible advances Accessibility.

There are clear indications that psychological theory has limited interoperability.
For one, theories are rarely \emph{``refuted nor corroborated, but instead merely fade away as people lose interest''} (Meehl, 1978).
To be interoperable, psychological theory would need to play a concrete part in scientists' day-to-day work.
For example, it should be possible to integrate theory directly into analysis workflows:
to derive hypotheses, select control variables, and guide model specification.
The aforementioned lack of formalization {[}REF Robinaugh{]} and ambiguity {[}REF Frankenhuis{]} prevent interoperability in such a practical sense.
Additionally, theories should be interoperable with each other:
for instance, it should be possible to embed a specific theory about the process of emotion regulation {[}REF Gross{]} within a theory of emotion regulation development {[}REF Morris{]}.

There appears to be a norm \emph{against} the reuse of theory,
as evident from the quip that
\emph{``{[}Theories are{]} like toothbrushes --- no self-respecting person wants to use anyone else's''} (Mischel, 2008).
Moreover, a legal basis for theory reuse is often absent.
Questions that often come up in our workshops on FAIR Theory are ``who owns theory'',
and ``who determines how a theory may be changed''?
Such questions require a clear answer, even if that answer may vary across theories.
Licensing theories for reuse provides clarity about how a theory can be reused.
A crucial legal consideration is that,
while specific (FAIR) representations of theory might be protected by copyright,
the underlying ideas are not.
This is another argument in favor of publishing specific instantiations of theory along with licenses detailing how they may be reused.

\subsection{FAIR Theory and Recognition \& Rewards}\label{fair-theory-and-recognition-rewards}

@@

\subsection{Version Control}\label{version-control}

\begin{itemize}
\tightlist
\item
  One source of potential improvements of theory methodology that has not been previously considered is computer science.
\item
  The process of ``iteratively improving'' digital objects - in this case, computer code - is well understood.
\item
  Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
\item
  This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
\item
  We introduce ``FAIR theory'', a digital research artifact to represent formal social scientific theories
\item
  FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner. Version control also enables diffuse collaboration in theory development, as other researchers can submit ``pull requests'' to suggest modifications of a theory, or can ``fork'' existing theories to create a spin-off from an existing theory.
\end{itemize}

@@

\subsection{Formalizing the Empirical Cycle}\label{formalizing-the-empirical-cycle}

In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
As several authors have taken inspiration from the work by De Groot,
we compare our interpretation of the original theory to the interpretation of others.
Originally, the theory has the following structure:

\begin{verbatim}

digraph {

  observation;
  induction;
  deduction;
  test;
  evaluation;

  observation -> induction;
  induction -> deduction;
  deduction -> test;
  test -> evaluation;
  evaluation -> observation;

}
\end{verbatim}

Subsequently, Wagenmakers and colleagues modified the theory by \emph{``{[}adding the{]} Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification''}:

\begin{verbatim}
digraph {

  subgraph cluster_discovery {
    label="Discovery";
    hypothesis [label="New hypothesis"];
    prediction [label="New prediction"];
  }
  data  [label="Old knowledge and old data"];      
  subgraph cluster_justification {
    label="Justification";
    test [label="Test on new data"];
    evaluation;
  }

  data -> hypothesis [label="Speculate & explore"];
  hypothesis -> prediction  [label="Deduce"];
  prediction -> test  [label="Design new experiment"];
  test -> evaluation  [label="Statistical analysis"];
  evaluation -> data  [label="Knowledge accumulation"];

}
\end{verbatim}

Note, however, that there appear to be further changes:
the phases of the cycle have been renamed,
and the annotations suggest a move towards experimental empirical psychology that was absent in the original formulation.
Moreover, the label ``knowledge accumulation'' invites the question of exactly \emph{how} knowledge accumulates upon evaluation of a prior experiment.
As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.

Our work, too is inspired by De Groot, but our take on the empirical cycle is different again:

\begin{verbatim}
digraph {

  theory;
  prediction;
  test [label="inferential procedure"];
  observation;

  theory -> prediction [label="deduction"];
  prediction -> test;
  test -> observation;
  observation -> theory [label="generalization"];

}
\end{verbatim}

In our representation,
induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory.
For example, the refutation of a hypothesized effect,
or the serendipitous observation of some pattern in data,
might be a reason to revise or construct theory.
Induction, incidentally, also occurs within the link from prediction to testing:
in the form of the inductive bias of methods used to perform the test,
and auxiliary assumptions that must be made to address remaining theoretical ambiguities.

@@

Some have argued that \emph{causal explanations} are a property of good theory {[}REF Meehl, etc?{]}.
According to Pearl and colleagues,
explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data.
Any formal theory that is explicit about direction of causality could thus be used to guide causal inference,
and could even be integrated into the analysis environment.

In this example, we illustrate how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG. We could use that to illustrate updating FAIR theory:

@@

\bibitem[\citeproctext]{ref-mischelToothbrushProblem2008}
Mischel, W. (2008). The {Toothbrush Problem}. \emph{APS Observer}, \emph{21}. Retrieved from \url{https://www.psychologicalscience.org/observer/the-toothbrush-problem}
